---
title: "Class12"
author: "Gabriella Tanoto (A18024184)"
toc: true
format: pdf
---

# Section 1 - Proportion of G/G in Population

```{r}
MXL <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```
> Q5. What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are homozygous for the asthma associated SNP (G|G)?

```{r}
table(MXL$Genotype..forward.strand.) / nrow(MXL)*100
```
9 out of 64 (14.01%) of the samples of Mexican ancestry individuals are homozygous G/G


Let's now compare it with the GBR population
```{r}
GBR <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Proportion of G|G in this GBR population is:
```{r}
round(table(GBR$Genotype..forward.strand.)/nrow(GBR)*100, 2)
```
Here, the proportion of homozygous G in the population is: 29.67%. It is a higher proportion — almost twice more homozygous GG than the MExican nacestry in LA.



# HOMEWORK CLASS 12
## Section 4: Population Scale Analysis

The following file will be used to determine if the SNP rs8067378 has any effect on the expression of ORMDL3 gene. It shows the genotype of the SNP and the expression level of ORMDL3. 

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 

Opening the file:
```{r}
snpfile <- read.table("rs8067378_ENSG00000172057.6.txt")
```

Genotypes sample size:
```{r}
table(snpfile$geno)
```

Median expression levels for each of the genotypes:
```{r}
#for Genotype A/A, the median expression level of ORMDL3 is 31.25
summary(snpfile[snpfile$geno=="A/A", ])
```
```{r}
#for Genotype A/G, the median expression level of ORMDL3 is 25.07
summary(snpfile[snpfile$geno=="A/G", ])
```
```{r}
#for Genotype G/G, the median expression level of ORMDL3 is 20.07
summary(snpfile[snpfile$geno=="G/G", ])
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? 

```{r}
library(ggplot2)

ggplot(snpfile) +
  aes(geno, exp, fill= geno)+
  ylab("Genotype") + xlab("Expression of ORMDL3") +
  scale_fill_manual(name= "Genotypes", values= c("#05ED99", "#0587ED", "#4149FB"))+
  ggtitle("Boxplot of ORMDL3 Expressions", 
          subtitle = "SNP rs8067378 genotypes are correlated with ORMDL3 Gene Expressions")+
  geom_boxplot(notch = T) + 
  geom_jitter(size=1.2, fill="darkgrey", alpha=0.3, width = 0.23) +
  theme_bw() +
  ggeasy::easy_center_title() 
  
```

According to this boxplot, the median expression levels of the ORMDL3 gene is quite different between the three SNP genotypes. ORMDL3 is most expressed when the individual is homologous A/A in their rs8067378 SNP. On the other hand, individuals with homologous G/G has the lowest expression of ORMDL3 gene. 